Topological Equivariant Artist Model for Visualization Library Architecture

### 2024-05-10 paper/main.tex
working bib + reworked rw + references + environment
proof of life? + equiv pics + start of rework of sheaf unpacking

% references
\usepackage[nocompress]{cite}
\usepackage[]{footmisc}
\usepackage{bookmark}

@@

@@

\title{Topological Equivariant Artist Model for Visualization Library Architecture}
% author names and IEEE memberships
\author{Hannah~Aizenman, Mikael Vejdemo-Johansson, Thomas~Caswell, and~Michael~Grossberg,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem H. Aizenman and M. Grossberg are with the department of Computer Science, City College of New York. 
\protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: haizenman@ccny.cuny.edu, mgrossberg@ccny.cuny.edu 
\IEEEcompsocthanksitem Mikael Vejdemo-Johansson is with the department of Mathematics, CUNY College of Staten Island.
\protect\\
E-mail: mvj@math.csi.cuny.edu
\IEEEcompsocthanksitem Thomas Caswell is with National Synchrotron Light Source II, Brookhaven National Lab 
\protect \\
E-mail: tcaswell@bnl.gov}% <-this % stops an unwanted space
\thanks{Manuscript received X XX, XXXX; revised X XX, XXXX.}
}

@@

@@

Since the earliest maps \cite{friendlyBriefHistoryData2006}, a visualization has been expected to preserve the $field$ properties and $topology$ of the corresponding dataset:
\begin{description}
  \item[\textcolor{fiber}{\textbf{field}} \footnote{Throughout this paper, we augment the math with color to group conceptually related terms\cite{headMathAugmentationHow2022}, for example \textcolor{fiber}{field} and \textcolor{fiber}{fiber}. This color coding carries through to the figures.}]  is a set of values of the same type, e.g. one column of a table or the pixels of an image
  \item[\textcolor{base}{\textbf{topology}}] is the connectivity and relative positioning of elements in a dataset \cite{wilkinsonGrammarGraphics2005}. 
\end{description}

@@

The structure on values has traditionally been codified using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale1971, thomasMathematizationNotMeasurement2014}.

Specifically, Steven's conceptualizes the structure on values as \textcolor{action}{actions} on groups \footnote{A \textit{group} is a set with an associative binary operator. This operation must have an identity element and be closed, associative, 
and invertable}:

@@

where the measurements are the groups and changes to them are actions. A function that preserves structure when the input or output is changed by a group action is called *equivariant*. Given a group $G$ that acts on both the input $X$ and the output $Y$ of a function $f: X \rightarrow Y$
\begin{definition}\label{def:equivariance}
 A function $f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$ \cite{pittsNominalSetsNames2013}
\end{definition}
which means that a visualization is structure preserving when there exist compatible group actions on the data and visualization, as proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}.

While *equivariance* allows us to test stucture preservation under transformations to data or visualization, *homomorphism* provides a way of asserting whether the relations between data elements was preserved as proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. Given the function $f: X \rightarrow Y$, with operators $(X, \circ)$ and $(Y, *)$

@@

\end{definition}
which means that the operators $\circ$ and $*$ are compatiable. For example, although the Steven's scales were conceptualized as having group structure, the ordinal scale have a monoidal structure because partial orders ($\geq, \leq$) are not invertable. \note{check grimalidi for citation}.

@@

Our framework for structure preservation incorporates equivariance and homomorphism and allows for defining other measures of equivalent representation; it is discussed in detail in \autoref{sec:artist:equivariant:data}.

\subsection{Topology}
Borrowing structures from algebraic topology allows us to describe structure on individual elements in a way that holds true for the entire dataset, meaning we can use the same tools to check if structre is preserved regardless of whether the data fits in memory, is distributed, or is streaming. Our framework generalizesis because, informally, a topology $\mathcal{T}$ is a partitioning of a space into its subspaces in a way where those subspaces can be composed back into the space; the choice of partitions is how we encode the connectivity of the data and visualizations that we are interested in.

@@

\begin{definition}\label{def:fiber_bundle}
   A \textbf{fiber bundle} $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and  bundle projection map $\pi: \dtotalc \rightarrow \dbasec$ \cite{FiberBundle2020,spanier1989algebraic}. 
\begin{equation}
  \label{eq:atct:fb:intro}
  \begin{tikzcd}[ampersand replacement=\&, row sep=huge]
   \dfiberc
    \arrow[r, hook, color=total] \& 
    \dtotalc
    \arrow[d, "\pi"',color=total, two heads] \\
     \& 
  \dbasec
     \arrow[u, "\dsectionc"', bend right, pos=.5, color=section, dashed]
  \end{tikzcd}
\end{equation} 
A continuous surjective map $\bm{\pi}$ is a \textbf{bundle projection} map when 
\begin{enumerate}
  \item all fibers in the bundle are isomorphic. Since all fibers are isomorphic, there is a uniquely determined \textcolor{fiber}{fiber space} \dfiberc\ given by the preimage of the projection $\pi$ at any point $\dbasepoint$ in the \textcolor{base}{base space} \dbasec: $\dfiberc = \pi^{-1}(k)$.  
  %the \textcolor{fiber}{fiber space} \dfiberc\ is the preimage of the projection function $\pi$ at a point $\dbasepoint$ in the \textcolor{base}{base space} \dbasec\ such that $\dfiber_{\dbasepoint} = \pi^{-1}(\dbasepoint)$. All fibers in a bundle are isomorphic such that $\dfiber \cong \dfiber_{\dbasepoint}$ for all points $\dbasepoint \in \dbase$. 
  \item each point $\dbasepointc$ in the \textcolor{base}{base space} \dbasec\ has an open neighborhood $\openset_{\dbasepointc}$ such that the \textcolor{total}{total space} \dtotalc\ over the neighborhood is locally trivial. \textbf{Local triviality} means $\dtotal\restriction_{\openset} = \openset\times F$. In this paper we use $\dtotal\restriction_{\openset} = \pi^{-1}(\openset)$ to denote the preimage of an openset, and a \textbf{local trivialization} is a specific choice of neighborhoods and their preimages.
  %there is an open neighborhood $\openset_{\dbasepoint}$ surrounding each point in the base space $\dbasepoint \in \openset_{\dbasepoint} \subset \dbase$ such that the \textcolor{total}{total space} \dtotalc\ over the neighborhood, denoted $\dtotal\restriction\openset$\footnote{The symbol $\restriction$ is the restriction operator\cite{RestrictionMathematics2022} defined in \texttt{amssymb}. For example $\pi^{-1}\restriction\openset$ is the inverse function $\pi^{-1}$ defined only on the points in $\openset$. Since $\pi^{-1}(\dbase)=\dtotal$, we use the shorthand $\dtotal\restriction_{\openset} \coloneqq \pi^{-1}\restriction_{\openset}$}, is locally trivial. The condition for local triviality is that $\dtotal\restriction_{\openset} = \openset \times \dfiber = \pi^{-1}\restriction_{\openset}$
\end{enumerate}
\end{definition}
\begin{definition} A \textcolor{section}{\textbf{section}} $\dsectionc: \dbasec \rightarrow \dtotalc$ over a fiber bundle is a smooth right inverse of $\pi$: $\pi(\dsection(\dbasepoint)) = \dbasepoint$ for all $\dbasepoint \in \dbase$
\end{definition}

\begin{figure}[H]
  \includegraphics[width=\columnwidth]{figures/local_trivalizations.png}
  \caption{The local trivializations $\dtotal_{0}$ and $\dtotal_{1}$ are subspaces of $\dtotal$ where the fibers $\dfiber_{0}$ are equal, here illustrated as all pointing in the same direction. These local trivializations are themselves locally trivial fiber bundles.} 
  \label{fig:atct:local_trivalizations}
\end{figure}

@@

This artist is then passed along to a shim artist that makes it compatible with existing Matplotlib objects (\autoref{sec:appendix:artist_shim}). This shim object is hooked into the Matplotlib draw tree to produce the vertical bar chart in \autoref{fig:discussion:vbar}. Using the Matplotlib artist framework means this new artist can be composed with existing artists, such as the ones that draw the axes and ticks. The example in this section is intentionally trivial to illustrate that the math to code translation is fairly straightforward and results in fairly self contained composable functions. A library applying these ideas, created by Thomas Caswell, can be found at \url{https://github.com/matplotlib/data-prototype}. Further research could investigate building new systems using this model, specifically libraries for visualizing domain specific structured data and domain specific artists. More research could also explore applying this model to visualizing high dimensional data, particularly building artists that take as input distributed data and artists that are concurrent. Developing complex systems could also be an avenue to codify how interactive techniques are expressed in this framework.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/https://www.overleaf.com/download/project/64063771726807fc06c1ee16/build/187faee483d-291434de4cddeda6/output/output.pdf?compileGroup=priority&clsiserverid=clsi-pre-emp-c2d-d-f-nkpv&enable_pdf_caching=true&popupDownload=truejjj
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/

\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{references}

>>>>

@@

% *** BIBLIOGRAPHY ***
\usepackage[]{footmisc}
\usepackage{bookmark}
\usepackage[style=ieee, backend=biber]{biblatex}
\addbibresource{references.bib}

@@

\title{Topological Equivariant Artist Model for Visualization Library Architecture}
% author names and IEEE memberships
\author{Hannah~Aizenman, Mikael Vejdemo-Johansson, Thomas~Caswell, and~Michael~Grossberg,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem H. Aizenman and M. Grossberg are with the department of Computer Science, City College of New York.
\protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: haizenman@.edu, mgrossberg@ccny.cuny.edu
\IEEEcompsocthanksitem Mikael Vejdemo-Johansson is with the department of Mathematics, CUNY College of Staten Island.
\protect\\
E-mail: mvj@math.csi.cuny.edu
\IEEEcompsocthanksitem Thomas Caswell is with National Synchrotron Light Source II, Brookhaven National Lab
\protect \\
E-mail: tcaswell@bnl.gov}% <-this % stops an unwanted space
\thanks{Manuscript received X XX, XXXX; revised X XX, XXXX.}
}

@@

Since the earliest maps \cite{friendlyBriefHistoryData2006}, a visualization has been expected to preserve the $field$ properties and $topology$ of the corresponding dataset:
\begin{LaTeXdescription}
  \item[\textcolor{fiber}{\textbf{field}} \footnote{Throughout this paper, we augment the math with color to group conceptually related terms\cite{headMathAugmentationHow2022}, for example \textcolor{fiber}{field} and \textcolor{fiber}{fiber}. This color coding carries through to the figures.}]  is a set of values of the same type, e.g. one column of a table or the pixels of an image
  \item[\textcolor{base}{\textbf{topology}}] is the connectivity and relative positioning of elements in a dataset \cite{wilkinsonGrammarGraphics2005}.
\end{LaTeXdescription}

@@

@@

The structure on values has traditionally been codified using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical structure\cite{stevensTheoryScalesMeasurement1946} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale1971, homasMathematizationNotMeasurement2014}.

 \textit{equivariance} allows us to test stucture preservation under transformations to data or visualization has been proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}.

 \textit{homomorphism} provides a way of asserting whether the relations between data elements was preserved as proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}.

\subsubsection{equivariance}
Specifically, Steven's conceptualizes the structure on values as \textcolor{action}{actions} on groups \footnote{A \textit{group} is a set with an associative binary operator. This operation must have an identity element and be closed, associative,
and invertable}:

@@

where the measurements are the groups and changes to them are actions. A function that preserves structure when the input or output is changed by a group action is called \textit{equivariant}. Given a group $G$ that acts on both the input $X$ and the output $Y$ of a function $f: X \rightarrow Y$
\begin{definition}\label{def:equivariance}
 A function $f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$ \cite{pittsNominalSetsNames2013}
\end{definition}
which means that a visualization is structure preserving when there exist compatible group actions on the data and visualization, as proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}.

\subsection{homomorphism}

For example, although the Steven's scales were conceptualized as having group structure, the ordinal scale have a monoidal structure because partial orders ($\geq, \leq$) are not invertable. This means we cannot use \textit{equivariance} to test for structure preservation instead we use \textit{homomorphsim} because it imposes fewer constraints on the underlying mathematical structure of the data.

Given the function $f: X \rightarrow Y$, with operators $(X, \circ)$ and $(Y, *)$

@@

\end{definition}
which means that the operators $\circ$ and $*$ are compatiable.  \note{check grimalidi for citation}.

@@

As shown in \autoref{fig:def:equivariance} and \autoref{fig:def:homomorphsim}, the basis of asserting structure preservation is that there exist compatiable actions and these actions commute; we generalize compatability and commutability as the basis for structure preservation in \autoref{sec:artist:equivariant:data}.

\subsection{Topology}
\note{lead with homomorphism}
Borrowing structures from algebraic topology allows us to describe structure on individual elements in a way that holds true for the entire dataset, meaning we can use the same tools to check if structre is preserved regardless of whether the data fits in memory, is distributed, or is streaming. Our framework generalizesis because, informally, a topology $\mathcal{T}$ is a partitioning of a space into its subspaces in a way where those subspaces can be composed back into the space; the choice of partitions is how we encode the connectivity of the data and visualizations that we are interested in.

@@

%\caption{The local trivializations $\dtotal_{0}$ and $\dtotal_{1}$ are subspaces of $\dtotal$ where the fibers $\dfiber_{0}$ are equal, here illustrated as all pointing in the same direction. These local trivializations are themselves locally trivial fiber bundles.}

\begin{definition}\label{def:fiber_bundle}
   A \textbf{fiber bundle} $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and  bundle projection map $\pi: \dtotalc \rightarrow \dbasec$ \cite{FiberBundle2020,spanier1989algebraic}.

    \begin{equation}
      \label{eq:atct:fb:intro}
      \begin{tikzcd}[ampersand replacement=\&, row sep=huge]
       \dfiberc
        \arrow[r, hook, color=total] \&
        \dtotalc
        \arrow[d, "\pi"',color=total, two heads] \\
         \&
      \dbasec
         \arrow[u, "\dsectionc"', bend right, pos=.5, color=section, dashed]
      \end{tikzcd}
    \end{equation}

A continuous surjective map $\bm{\pi}$ is a \textbf{bundle projection} map when
\begin{enumerate}
  \item all fibers in the bundle are isomorphic. Since all fibers are isomorphic, there is a uniquely determined \textcolor{fiber}{fiber space} \dfiberc\ given by the preimage of the projection $\pi$ at any point $\dbasepoint$ in the \textcolor{base}{base space} \dbasec: $\dfiberc = \pi^{-1}(k)$.
  %the \textcolor{fiber}{fiber space} \dfiberc\ is the preimage of the projection function $\pi$ at a point $\dbasepoint$ in the \textcolor{base}{base space} \dbasec\ such that $\dfiber_{\dbasepoint} = \pi^{-1}(\dbasepoint)$. All fibers in a bundle are isomorphic such that $\dfiber \cong \dfiber_{\dbasepoint}$ for all points $\dbasepoint \in \dbase$.
  \item each point $\dbasepointc$ in the \textcolor{base}{base space} \dbasec\ has an open neighborhood $\openset_{\dbasepointc}$ such that the \textcolor{total}{total space} \dtotalc\ over the neighborhood is locally trivial. \textbf{Local triviality} means $\dtotal\restriction_{\openset} = \openset\times F$. In this paper we use $\dtotal\restriction_{\openset} = \pi^{-1}(\openset)$ to denote the preimage of an openset, and a \textbf{local trivialization} is a specific choice of neighborhoods and their preimages.
  %there is an open neighborhood $\openset_{\dbasepoint}$ surrounding each point in the base space $\dbasepoint \in \openset_{\dbasepoint} \subset \dbase$ such that the \textcolor{total}{total space} \dtotalc\ over the neighborhood, denoted $\dtotal\restriction\openset$\footnote{The symbol $\restriction$ is the restriction operator\cite{RestrictionMathematics2022} defined in \texttt{amssymb}. For example $\pi^{-1}\restriction\openset$ is the inverse function $\pi^{-1}$ defined only on the points in $\openset$. Since $\pi^{-1}(\dbase)=\dtotal$, we use the shorthand $\dtotal\restriction_{\openset} \coloneqq \pi^{-1}\restriction_{\openset}$}, is locally trivial. The condition for local triviality is that $\dtotal\restriction_{\openset} = \openset \times \dfiber = \pi^{-1}\restriction_{\openset}$
\end{enumerate}
\end{definition}
\begin{definition} A \textcolor{section}{\textbf{section}} $\dsectionc: \dbasec \rightarrow \dtotalc$ over a fiber bundle is a smooth right inverse of $\pi$: $\pi(\dsection(\dbasepoint)) = \dbasepoint$ for all $\dbasepoint \in \dbase$
\end{definition}

\begin{figure}[H]
    \includegraphics[width=\columnwidth]{figures/local_trivalizations.png}
    \label{fig:atct:local_trivalizations}
    \caption{Example of bundles $\dtotalc_0$ and $\dtotalc_1$, which are both trivializations
    of a bundle $\dtotalc$ add in section here!}
\end{figure}

\autoref{fig:atct:local_trivalizations} shows an example of a two dimenensional bundle
$\dtotal$. This bundle can encapsulate a set datasets where the data points are connected to each other along a unit circle $\dbasec$ = $\left[0, 2\pi\right]$ and the data values lie along an interval $dfiberc$ = $[0,1]$. Each dataset is encoded as a section, for example $sin$ and $cos$. The subspaces of $\dtotal_{0}$ and $\dtotal_{1}$ are local trivializations $\dtotal$; therefore the fibers are equal, here illustrated as all pointing in the same direction. These local trivializations are themselves locally trivial fiber bundles.

@@

This artist is then passed along to a shim artist that makes it compatible with existing Matplotlib objects (\autoref{sec:appendix:artist_shim}). This shim object is hooked into the Matplotlib draw tree to produce the vertical bar chart in \autoref{fig:discussion:vbar}. Using the Matplotlib artist framework means this new artist can be composed with existing artists, such as the ones that draw the axes and ticks. The example in this section is intentionally trivial to illustrate that the math to code translation is fairly straightforward and results in fairly self contained composable functions. A library applying these ideas, created by Thomas Caswell and Kyle Sunden, can be found at \url{https://github.com/matplotlib/data-prototype}. Further research could investigate building new systems using this model, specifically libraries for visualizing domain specific structured data and domain specific artists. More research could also explore applying this model to visualizing high dimensional data, particularly building artists that take as input distributed data and artists that are concurrent. Developing complex systems could also be an avenue to codify how interactive techniques are expressed in this framework.

@@

\printbibliography
